---
title: "Belgian Grid Forecast"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    theme: united
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(flexdashboard)
library(dplyr)
library(ggplot2)
library(plotly)
library(DT)
library(lubridate)

data <- readr::read_csv('data/results.csv')
```

# Forecast

## Column {data-width="650"}

Last refresh: `r Sys.time()`

### 3 day ahead forecast

```{r}
p1 <- data %>% 
  filter(type == 'preds') %>% 
  mutate(Predicted = round(.pred, 2)) %>% 
  ggplot(aes(x=datetime, y = .pred, label = Predicted))+
  geom_point(size=0.5)+
  geom_line()+
  theme_bw()+
  ylab('Predicted Load -  MW')+
  xlab(NULL)

ggplotly(p1, tooltip=c("Predicted", 'datetime'))
```

## Column {data-width="350"}

### Past 1 day backfit

The final day is held back from the model to assess performance on unseen data for the forecast.

```{r}
p2 <- data %>% 
  filter(type == 'backfit') %>% 
  ggplot()+
  geom_line(aes(x=date_time, y = total_load), alpha=0.5)+
  geom_line(aes(x=date_time, y = .pred), color='blue', alpha=0.5)+
  theme_bw()+
  ylab('Load -  MW')+
  xlab(NULL)+
  labs(title = 'Predicted values (blue) vs. Real Load (black)')

ggplotly(p2)
```

### Status

```{r}
latest <- data %>% 
  filter(type != 'preds') %>% 
  pull(date_time) %>% 
  max()-days(1)

backtest <- data %>% 
  filter(type != 'preds') %>% 
  pull(date_time) %>% 
  max() %>% 
  date()

forecast_from <- data %>% 
  filter(type == 'preds') %>% 
  pull(date_time) %>% 
  min()

mae_calc <- data %>% 
  filter(type != 'preds') %>%
  yardstick::mae(.pred, total_load) %>% 
  pull(.estimate)

cat(paste0('Last data model was trained on: ', latest, '.'))
cat(paste0('Back test prediction for: ', backtest))
cat(paste0('Forecast from: ', forecast_from))
cat(paste0('Seen Mean Absolute Error (per period): ', round(mae_calc, 2), ' MW'))
```

# Table View

## Column {data-width="350"}

### 3 day ahead forecast

This is a tabular view of the main graph.

```{r}
data %>% 
  filter(type == 'preds') %>% 
  select(datetime, .pred, average_temp) %>% 
  DT::datatable(extensions = 'Buttons', 
                options = list(
    dom = 'Bfrtip',
    searchable = FALSE,
    buttons = c('copy', 'print', 'csv', 'excel')))%>% 
  formatRound('.pred', 0) %>% 
  formatDate('datetime', method='toUTCString')
```

# About This

### What is this

This is a quick and dirty experiment using Belgium's largest grid operator (Elia)'s [generous open data commitment](https://opendata.elia.be/pages/home/) to model the previous load and forecast ahead a 3 day horizon.

To check with the actual grid situation, [visit here](https://www.elia.be/en/grid-data) and scroll down to the grid load chart.

### How does this work?

Every day at midnight UTC a script automatically downloads the previous day's load from the elia website, as well as the temperatures in Brussels and Bruges from the [openmeteo API.](https://open-meteo.com/)

A machine learning technique called gradient boosting is used to make a time series prediction for the next 3 days.

A number of different features are included in the final model, including holiday and weekend times, warming effects and recent loads.

A back fit on the past day is also run as a quick a validation step. This allows a Mean Absolute Error calculation: the average error expected, either + or - per 15 minute period.

For some exploratory data analysis I sketched up while working on this model, [see here.](https://charlesmercieca.github.io/EDA)

[www.charlesmercieca.com](www.charlesmercieca.com)
